---
title: "Very basic statistical computing in R"
author: "Malachy Campbell"
date: "11/9/2018"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Outline
- Very basics of R
    - Types of objects and classes
    - Input and output (setwd, read/write)

- Manipulating data 
    - manipulating datasets with reshape2 and tidyr
    - Applying functions over datasets (for loops, apply and plyr)
    
- Data visualization
    - Base graphics and ggplot2)

- Presenting data
    - Reproducible reports and presentations in Rmarkdown


## Basics of R: Objects and object types

- Programming in R is centered around the use of objects (data structures with attributes)

    - Apply some method or function on object
    
- These objects can be of several different types (character, factor, dataframe, list, etc.)

\tiny
```{r objects and vectors, echo = T}
someObject <- 5

#What type of object is this?
str(someObject)
is.numeric(someObject)

anotherObject <- "5"
str(anotherObject)
is.character(anotherObject)

#Must quote strings or characters or else they are treated as objects
finalObject <- "five"
str(finalObject)
is.factor(finalObject)
```

## Basics of R: Manipulating objects
- The object type will restrict the functions that can be used on the object

\tiny
```{r manipulate objects, echo = T}
#Try to add some number to 'anotherObject', a character vector containing '5'
try(anotherObject + 5)

#as.numeric forces some object to be numeric
anotherObject <- as.numeric(anotherObject)
anotherObject + 5

#finalObject contains the string "five". Try and convert "five" to some number. R doesn't like this.
as.numeric(finalObject)
```

## Basics of R: Vectors, matrices, data frames, and lists
- All of these store more than one element

\tiny
```{r vector matrix df list, echo = T}
#c() combines 
aVector <- c(anotherObject, 16, 20, 22, 27, 30)
aVector
mean(aVector)

#Access the 5th element of aVector
aVector[5]

aMatrix <- matrix(aVector, nrow = 2, ncol = 3, byrow = T)
aMatrix

#Access the element in the second row and third column of aMatrix [x,y]
aMatrix[2,3]
```

## Basics of R: Data frames
- A data frame is an object that can handle multiple types of data

\tiny
```{r df, echo = T}
#Rep(X, N): repeat some value(s) N times 
aDataframe <- data.frame(Obs = rep(c("A", "B"), 3), Trt = rep(c("T1", "T2"), 3), Val = aVector)
str(aDataframe)

#show only the first three rows. Rows are indicated by [x,], columns are [,x]
aDataframe[1:3,]

#access specific columns using object$'column name' or object[,colnumber] 
aDataframe$Obs[1:3]
mean(aDataframe$Val)
```

## Basics of R: Lists
- Lists can store any type of data in each of its elements

\tiny
```{r list, echo = T}
firstList <- list(aVector, aMatrix, aDataframe)
firstList[1]

nestedList <- list(c(1,2,3,4), firstList)
nestedList
```

## Input and output
- R can read various types of text files (.csv, .txt, .xlsx (try to avoid), etc.)

\tiny
```{r input and output, echo = T}
setwd("/Users/malachycampbell/Documents/Dropbox/Work/Presentations/Japan/IntroToR/")
#Data <- read.csv("/Users/malachycampbell/Documents/Dropbox/Work/Presentations/Japan/IntroToR/SeriousData.csv")
Data <- read.csv("SeriousData.csv")
head(Data)
str(Data)

write.csv(Data, "SomeData.csv", row.names = F)
```

## Input and output
- R can read various types of text files (.csv, .txt, .xlsx (try to avoid), etc.)

\tiny
```{r input and output2, echo = T}
str(Data)

write.csv(Data, "SomeData.csv", row.names = F)
```

## Cleaning up data
\tiny
```{bash show ugly data, echo = F}
head /Users/malachycampbell/Documents/Dropbox/Work/Presentations/Japan/IntroToR/SillyData.csv
```

\tiny
```{r input and output ugly data, echo = T}
Data <- read.csv("SillyData.csv")
head(Data)
```

## Cleaning up data
\tiny
```{r input and output ugly data2, echo = T}
Data <- read.csv("SillyData.csv", header = T, skip = 2)
str(Data)

#Replace 'Missing' with NA
Data[Data$Height == "Missing", ]$Height <- NA

#Convert the height column to numeric
Data$Height <- as.numeric(as.character(Data$Height) )
```

## Cleaning up data
\tiny
```{r input and output ugly data3, echo = T}
head(Data)
str(Data)
```