---
title: "R markdown report"
author: "Malachy Campbell"
date: "`r Sys.Date()`"
output:
  rmdformats::html_clean:
    fig_width: 6
    fig_height: 6
    highlight: kate
    thumbnails: true
    lightbox: true
    gallery: true
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '~/Desktop/')
```

# Background
The purpose of this is to show everyone how markdown works. The first header is called "Background".

## This is a subheader
Here, I'll show how to change the font style and create an itemized list.

* One asterisk for *italic*
  - Subitem
* Two for **bold**
* tildes for ~~strike through~~.

### Code chunks
#### R code
To specify a chunk of code start with ```{engine, evaluate code?, show code?...} and end with ```. This is a chunk of R code. The engine is R, we will run the code, and show the output. 
```{r code example, eval = T, echo=T}
HiR <- "Hi from R!!"

print(HiR)
```

#### bash code
We can also call bash codes from inside our markdown document. Here the engine is bash, so we change the first line accordingly. 
```{bash hi example, echo=T}
echo 'Hi, its me again!'
```

```{bash quiet hi example, eval=F, echo=T}
echo 'Hi, its me again, but this time there no output :('
```

# Show some plots
This is a plot aligned to the left.
```{r plot hist, echo=T, fig.align='left', fig.height=4, fig.width=3}
hist(rnorm(100, 10, 2), breaks = 20)
```

Here's one to the right.
```{r plot hist2, echo=T, fig.align='right', fig.height=3, fig.width=4}
hist(rnorm(100, 10, 2), breaks = 20)
```

# Displaying formulas using latex
Enclosing some latex code in one \$ will produce an inline formula $y = x + e$. If you use two \$ it will be placed beneath the text $$ y = x + e $$

Online equation builders can be helpful for beginners check one out [here](https://www.codecogs.com/latex/eqneditor.php). 

# More resources for learning R
Here are a few links for users that are new to R.
* [R for cats](https://rforcats.net/)
* [An Introduction to R](https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf): A comprehensive guide to using R.
* [R for data science](https://r4ds.had.co.nz)
* [Stack overflow](https://stackoverflow.com/questions/tagged/r) : A place to ask questions, and find answers