---
title: "Lab: Applications for Ordinary Least Squares and Mixed Models"
author: "Malachy Campbell"
date: "10/16/2018"
header-includes:
   - \usepackage{bbm}
   - \usepackage{amsmath}
output: 
  beamer_presentation:
  theme: "CambridgeUS"
  colortheme: "beaver"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, size = "footnotesize")
library(lsmeans)
library(lme4)
```


# OLS and MM example 1: Balanced maize data
Learning objectives:

  1. Brief overview of ordinary least squares (OLS) and mixed models (MM)

  2. Estimate genetic values and $h^2$ using OLS and MM. 

  3. Learn to deal with unbalanced data
    
# Maize Dataset
- 62 recombinant inbred line (RILs) from a cross between B73 and MO17. 
  
- Randomized complete block design
  
- Two replications at four locations
  
- Traits: days to pollen, days to silking, anthesis/silking interval (ASI) and plant height.
  
    - We'll use height as the response variable.

# Loading the data.

\footnotesize
```{r load data, echo = T}
maize <- read.csv("~/Downloads/MaizeRILs.csv")

head(maize)
```

# Obtaining genetic values with OLS
- For this dataset we can fit the following model:

$$y_{ijk} = \mu + L_i + Rep(L)_{ij} + G_k + GL_{ik} + e_{ijk}$$ 

- $y_{ijk}$ is the phenotype (height)
    
- $L_i$ is the fixed effect of location $i$
    
- $Rep(L)_{ij}$ is the fixed effect of replicate $j$ nested within location $i$
    
- $G_k$ is the fixed effect of RIL $k$, $GL_{ik}$ is the interaction of RIL $k$ and location $i$ and $e_{ijk}$ is the residual.

**Here's everything except the error term is considered as a fixed effect**

# Obtaining genetic values with OLS

- Fit the linear model with lm in R

\footnotesize
```{r run anova for maize, echo = T, eval = T}
#rep is coded as 1 and 2. So make sure R knows its a factor
maize$rep <- as.factor(maize$rep)
mod1 <- lm(height ~ location*RIL + rep:location, data = maize)

#anova(mod1)
```

# Obtaining genetic values with OLS

-  Use the output of lm to estimate the marginal means

-  For RIL-11 we can calculate the marginal means as:

$$RIL11 = \mu + \bar{L} + G_{RIL11} + \bar{GL}_{RIL11} + \bar{Rep(L)}$$

# Obtaining genetic values with OLS
\footnotesize
```{r marginal means for RIL11, echo = T}
#intercept
MU <- as.numeric(coef(mod1)["(Intercept)"] )
#locations
LOC.eff <- sum(as.numeric(coef(mod1)[c("locationCLY",
          "locationPPAC","locationTPAC")] ))/4
#RIL
RIL1.eff <- as.numeric(coef(mod1)["RILRIL-11"] )
#RIL x Location
RIL1.LOC.eff <- sum(as.numeric(coef(mod1)
              [c("locationCLY:RILRIL-11",
                 "locationPPAC:RILRIL-11",
                 "locationTPAC:RILRIL-11")] ))/4
#Rep within location
Rep.eff <- sum(as.numeric(coef(mod1)[c("locationARC:rep2", 
              "locationCLY:rep2", "locationPPAC:rep2",
              "locationTPAC:rep2")] ))/8

RIL_11 <- MU + LOC.eff + RIL1.eff + RIL1.LOC.eff + Rep.eff

print(RIL_11)
```

# Estimating heritability from ANOVA/OLS
- Since the design is balanced we can estimate $H^2$ using ANOVA

$$h^2 = \frac{ \sigma_{RIL}^2 }{ \sigma_{RIL}^2 + \frac{\sigma_{RIL \times LOC}^2}{n_l} + \frac{\sigma_e^2}{n_r n_l} }$$.

-  EMS from the ANOVA table:

    $\sigma_{RIL \times LOC}^2 = \frac{MS(RIL \times LOC) - MS(Error)}{n_r}$, $\sigma_{RIL}^2 = \frac{MS(RIL) - MS(RIL \times LOC)}{n_r n_l}$, and  
$\sigma_e^2 = MS(Error)$

# Estimating heritability from ANOVA/OLS
```{r H2 for maize height, echo = T}
anova.res <- as.data.frame(anova(mod1))

sigma_err <- anova.res[5,3]
sigma_G.E <- (anova.res[3,3] - sigma_err) / 2  
sigma_G <- (anova.res[2,3] - anova.res[3,3]) / 8

H2.OLS <- sigma_G / (sigma_G + sigma_G.E/4 + sigma_err/8)
print(H2.OLS)
```

# Obtaining genetic values (BLUEs) with a mixed model
- We will fit a mixed model to estimate line values for each RIL 

    - $RIL$ as a fixed effect, and $Loc$ and $Rep$ as random effects
    
    - $Var(Loc) \sim N(0, \mathbf{I}\sigma^2_{LOC})$, $Var(rep) \sim N(0, \mathbf{I}\sigma^2_{rep})$, and $Var(e) \sim N(0, \mathbf{I}\sigma^2_{e})$

# Obtaining genetic values (BLUEs) with a mixed model in lme4
- Random terms are specified by '(1|some term)'.
  - '(1|location/rep)' is the random effect of rep nested within location
  - '(1|location:RIL)' is the random effect of location x RIL interaction

```{r MM BLUEs for maize height, echo = T}
library(lme4)
mod2 <- lmer(height ~ RIL + (1|location/rep) + (1|location:RIL), maize)

#List the estimates for the fixed effects
summary(mod2)$coefficients[1,1] + summary(mod2)$coefficients[2,1]
```

# Estimating heritability with a mixed model in lme4
- Here, all terms with the exception of $\mu$ will be considered random. 

```{r MM H2 maize height, echo = T}
mod3 <- lmer(height ~ 1 + (1|RIL) + 
               (1|location/rep) + 
               (1|location:RIL), maize)

#extract the variance components
MM.varcomps <- as.data.frame(VarCorr(mod3))

sigma_err.MM <- MM.varcomps[5,4]
sigma_G.E.MM <- MM.varcomps[1,4] 
sigma_G.MM <- MM.varcomps[2,4]

H2.MM <- sigma_G.MM / 
  (sigma_G.MM + sigma_G.E.MM/4 + sigma_err.MM/8)

print(H2.MM)
```

# BLUPs for maize height
- When we want to make a prediction on a random term in the model the predicted value is called BLUP

- In lme4:

```{r MM blups maize height, echo = T}
mod3 <- lmer(height ~ 1 + (1|RIL) + (1|location/rep) + (1|location:RIL), maize)

#extract the blups for RILs
blups_m3 <- ranef(mod3)$RIL
```

- More on BLUPs later!

# On your own
- Run a similar analyis with the unbalanced data and compare OLS and MM approaches

- Which is more trustworthy?